---
title: "ST332 Project 2"
author: "Group a"
date: "2/9/2021"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE)
```

```{r , echo=FALSE, message=FALSE} 
library(knitr) #wanna use template here
library(finalfit)
library(naniar)
library(visdat)
library(ggplot2)
library(tidyr)
library(tidyverse)
library(GGally)
library(gsynth) #causal inference with interactive fixed-effect models
library(MASS)
library(fitdistrplus)
library(logspline)
library(mice)
```

## Intro


## Data Collection
This report uses data collected in the CAST study (detailed by Lamb et.al., 2009) on 585 patients with ankle sprains. The study analyses different treatments comparatively, with the aim of making clinical recommendations. Encompassing 238 females and 347 males of ages 16-72, the dataset contains variables describing the age, weight, sex and height of each participant, as well as five health scores, which are components of the Foot and Ankle Outcome Score in the initial paper.

```{r import}
cast <- read.csv("CAST2miss.csv")
cast <-as.data.frame(cast)
```
```{r summ, echo=TRUE, output=TRUE}
summary(cast)
```


## Data entry check
The dataset used in the subsequent analysis is imported from a medical trial. We want to find out if the data entry process is robust by checking for variable type (coding) and looking at a summary table of the data.
```{r , echo=TRUE, output=TRUE}
ff_glimpse(cast)
```
It seems that the CAST dataset encompasses categorical variables and continuous variables. Sex is coded as integer despite being a factor, while the baseline sports variable is also coded as an integer, despite being a continuous variable. We convert sex into a factor with two levels, 0 (female) and 1 (male), and the baseline sports sub-scale into a continuous variable.

```{r ,include=FALSE}
castsp <- cast[, c(1:4,8,13)]
castsp %>% mutate(sex=sex-1)
castsp$sex <- as.factor(castsp$sex)
```

The four basic variables (age, sex, height, weight) are recorded for all patients, while the baseline sub-scale measurements are each missing in 1-2 cases, with the exception of the sport&recreation sub-scale which is missing in 12 cases. The subsequent measurements (after 9 months) for pain score, quality of life score and other symptoms score are also missing in lower numbers (all missing in 135 cases) than the daily life score and sports score, which are missing in 214, and, respectively, 218 cases.


## General plots 
  
We begin with some exploratory plots to visualize missingness in the dataset.
```{r missing viz, output=TRUE, echo=TRUE}
gg_miss_var(cast,show_pct=TRUE)

vis_miss(cast)
gg_miss_upset(cast)
```
From the figures above, it is noteworthy that missingness is most prevalent among the 9-month measurements of the sub-scales except other symptoms of pain variable, and the composite score, as follows:
* around 12% of the values are missing from the dataset (Fig 1)
* around 1-2% of the baseline sub-score values are missing (Fig 1)
* in 135 cases, all sub-scale variables are missing at 9-month point (Fig 2)
* in 78 cases, both daily living score at 9 months and sport & recreation score at 9 months are missing, so there might be a correlation between their missingness patterns (Fig 2)
* there are 5 cases where  only sports & recreation sub-scale values at 9 months are missing (Fig 2)
* there is 1 case where only the daily living scale at 9 months (Fig 2).


## General patterns and relationships

The function in sport and recreation variables (bsport, sport9) are missing 2.12% and 38.58% of the time. From Fig 1, it is clear that all 11 variables in question have some degree of missingness so we change the default settings of the visualization function to account for all interactions between all the missing variables (nintersects=NA).

```{r , output=TRUE, echo=TRUE}
gg_miss_upset(cast, nsets=16,nintersects=NA)
```

Fig above reveals more patterns of missingness:
* there are 130 cases where all sub-scales and the composite score are missing
* there are 4 cases when baseline sports & recreation and all other sub-scales at 9 months are missing, but 3 cases when only baseline sports & recreation is missing and 2 cases when only sports & recreation at 9 months is missing
* there are 75 cases where sports & recreation and daily living sub-scales are missing
* interestingly, across missingness counts, sports variable seems to be missing when daily living variable is missing.


```{r }
md.pattern(cast)
```


## Sports & recreation variable

Let's explore a summary of the CAST data with the basic variables and the sports sub-scale.



```{r }
md.pattern(castsp)
```




There are 304 full observations, 209 observations which are only missing the value at the 9-month time point and 3 observations which are missing the baseline value. The remaining 9 observations are missing both values. The former two categories will be of particular interest to the following analysis.

Let's now look at a plot showing the relationship between the prevalence of the sports variable at the two time points- baseline and 9 months.

```{r }
ggplot(cast,
       aes(x = bsport,
           y = sport9)) +
 geom_miss_point()
```


Let's explore the missingness relationship between baseline and 9-month measurements of the sports sub-scale, dependent on sex.

```{r }
ggplot(cast,
       aes(x = bsport,
           y = sport9))+ geom_miss_point()+facet_wrap(~sex)
```




Let's look at a plot showing the missingness of the CAST dataset variables by sex.



```{r }
gg_miss_fct(x = cast, fct = sex)
```

There seems to be a distinct correlation between missingness of sub-scale values (9-month measurement) and sex, with scores for females missing more sparsely than those for males. At baseline, there does not seem to be a significant difference in willingness to complete the survey between sexes.



## Missingness mechanism

We have to decide whether the missing 9-month values of the sports sub-scale are MCAR, MAR or MNAR.  

### MCAR vs MAR

If the sub-scale variables are MCAR, list-wise deletion would not introduce bias in our models and subsequent inferences. Let's use Little's test [1] to diagnose whether there are any variables missing completely at random in our dataset. We will also create dummy variables for missingness (1 = missing, 0 = observed).

If the sub-scale variables are MAR, their missingness is conditional on other variables, and should therefore be analyzed further.

```{r }
library(norm)
```
```{r mcar function}
mcar <- function(x){ 
    if(!require(norm)) {
        stop("You must have norm installed to use LittleMCAR") 
    } 

    # if(!require(data.table)) {
    #   stop("Please install the R-package data.table to use mcar")
    # }

    if(!(is.matrix(x) | is.data.frame(x))) {
        stop("Data should be a matrix or dataframe")
    }

    if (is.data.frame(x)){
        x <- data.matrix(x)
    }

    # delete rows of complete missingness
    foo <- function(x) return(any(!is.na(x)))
    dd <- apply(X = x, MARGIN = 1L, FUN = foo)
    dd <- which(!dd, arr.ind = TRUE)
    if(length(dd) > 0) 
        x <- x[-dd,]

    # define variables        
    n.var <- ncol(x) # number of variables
    n <- nrow(x)  #number of respondents
    var.names <- colnames(x)
    r <- 1 * is.na(x)

    nmis <- as.integer(apply(r, 2, sum))  #number of missing data for each variable REWRITE
    mdp <- (r %*% (2^((1:n.var - 1)))) + 1  #missing data patterns
    x.mp <- data.frame(cbind(x,mdp)) # add column indicating pattern
    colnames(x.mp) <- c(var.names,"MisPat") # set name of new column to MisPat
    n.mis.pat <- length(unique(x.mp$MisPat)) # number of missing data patterns
    p <- n.mis.pat-1 # number of Missing Data patterns minus 1 (complete data row)


    s <- prelim.norm(x)
    ll <- em.norm(s)
    fit <- getparam.norm(s = s, theta = ll)

    # gmean<-mlest(x)$muhat #ML estimate of grand mean (assumes Normal dist)
    gmean <- fit$mu
    # gcov<-mlest(x)$sigmahat #ML estimate of grand covariance (assumes Normal dist)
    gcov <- fit$sigma
    colnames(gcov) <- rownames(gcov) <- colnames(x)

    #recode MisPat variable to go from 1 through n.mis.pat
    x.mp$MisPat2 <- rep(NA,n)
    for (i in 1:n.mis.pat){ 
        x.mp$MisPat2[x.mp$MisPat == sort(unique(x.mp$MisPat), partial=(i))[i]]<- i 
    }

    x.mp$MisPat<-x.mp$MisPat2
    x.mp<-x.mp[ , -which(names(x.mp) %in% "MisPat2")]

    #make list of datasets for each pattern of missing data
    datasets <- list() 
    for (i in 1:n.mis.pat){
        datasets[[paste("DataSet",i,sep="")]]<-x.mp[which(x.mp$MisPat==i),1:n.var]
    }

    #degrees of freedom
    kj<-0
    for (i in 1:n.mis.pat){ 
        no.na<-as.matrix(1* !is.na(colSums(datasets[[i]]))) 
        kj<-kj+colSums(no.na) 
    }

    df<-kj -n.var

    #Little's chi-square
    d2<-0
    cat("this could take a while")

    # this crashes at the missingness pattern where every column is missing
    # this for-loop can be handled faster with plyr-function
    for (i in 1:n.mis.pat){ 
        mean <- (colMeans(datasets[[i]])-gmean) 
        mean <- mean[!is.na(mean)] 
        keep <- 1* !is.na(colSums(datasets[[i]])) 
        keep <- keep[which(keep[1:n.var]!=0)] 
        cov <- gcov 
        cov <- cov[which(rownames(cov) %in% names(keep)) , which(colnames(cov) %in% names(keep))] 
        d2 <- as.numeric(d2+(sum(x.mp$MisPat==i)*(t(mean)%*%solve(cov)%*%mean)))
    }

    #p-value for chi-square
    p.value<-1-pchisq(d2,df)

    #descriptives of missing data
    amount.missing <- matrix(nmis, 1, length(nmis))
    percent.missing <- amount.missing/n
    amount.missing <- rbind(amount.missing,percent.missing)
    colnames(amount.missing) <- var.names
    rownames(amount.missing) <- c("Number Missing", "Percent Missing")

    list(chi.square = d2, 
         df = df, 
         p.value = p.value, 
         missing.patterns = n.mis.pat, 
         amount.missing = amount.missing, 
         data = datasets)
}
```

```{r mcar, echo=FALSE}
r <- mcar(cast)
```

```{r mcar vs mar call, echo=TRUE, include=TRUE}
r[["p.value"]]
r[["missing patterns"]]
r[["amount.missing"]]
```




The output of Little's MCAR test indicates that no variables are MCAR.



```{r pairs, message=FALSE, warning=FALSE}
explanatory = c("age", "sex", 
  "height", "weight",  
  "bsport")
dependent = "sport9"

castsp %>% 
  missing_pairs(dependent, explanatory, position = "fill", )

castsp %>% 
  missing_compare(dependent, explanatory)
```
```{r }
explanatory = c("age", "sex", 
  "height", "weight")
dependent = "bsport"
castsp %>% 
  missing_compare(dependent, explanatory)
```


The output indicates a signficant relationship between age and the missingness of the 9-month sport sub-scale measurement, and a weak relationship between the two sport sub-scale variables. This result, coupled with Little's test, allows us to infer that bsport seems to be MCAR with respect to the basic variables, while sport9 is at least MAR. 

### Missingness Solution for Baseline Values

Because "bsport" is MCAR, the missing values can be completed by using multiple imputation. Prior to this, we perform sensitivity analysis because this variable is very important to the subsequent statistical modelling. There are only 12 observations missing the bsport value, which accounts for approximately 2% of the observations. 
  

```{r }
library(fastDummies)
castsp<- castsp %>% mutate(bsport_miss=bsport-bsport, sport9_miss=sport9-sport9)

castsp <- dummy_cols(castsp, select_columns = c("bsport_miss","sport9_miss"), remove_first_dummy=TRUE)

castsp <- castsp[,-c(7,8)]

castsp <- castsp %>% 
  rename(
    bsportna= bsport_miss_NA,
    sport9na=sport9_miss_NA
    )
```

```{r mice design, include=FALSE}
castsp_m1 <- mice(castsp,m=5,maxit=50,meth='pmm')
```
```{r mice, output=FALSE, echo=FALSE}
castsp_complete<-complete(castsp_m1,1)
```
```{r }
xyplot(castsp_m1,sport9 ~ height+age+sex+bsport,pch=18,cex=1)
```


### MNAR
After concluding that sport9 is at least MAR, let's test whether it can be MNAR. Understanding the missingness mechanism guides the statistical modelling technique.

```{r mar vs mnar, warning=FALSE, message=FALSE}

```

## Modelling missingness

We will use covariance analysis adjusted for age, sex, height and the appropriate baseline variable to establish the size and statistical significance of clinical differences at the 9-month timepoint.

## Modelling sport9 variable

```{r }
descdist(castsp$sport9na) 
``` 


The Cullen-Frey graph for sport9 indicates that its distribution resembles the uniform and Beta distributions.



## Limitations
* (bsport) List-wise deletion of the observations in which bsport is missing introduces a bias in the inferences. There is a weak relationship between bsport and age, which means that list-wise deletion implies keeping the observations from younger people, on average. This may have consequences for our conclusions if age is associated with sport9.

* (Yscore) List-wise deletion of observations leads to a bigger drop in male than female participants (as the score is proportionately missing), which might affect the inferences

* (reducing to no na's) Reduction of the dataset to participants with no missing data introduces a higher bias in the results. Apart from subsetting participants with no missing observations for the baseline scores, which account for a fairly small percentage of the dataset (approx. 12%), we would also drop all participants with at least one missing sub-scale, which account for 40% of the dataset. 

## References

1. Little, R. J. A. (1988). A test of missing completely at random for multivariate data with missing values. Journal of the American Statistical Association, 83(404), 1198--1202.

2. Jamshidian, M. Jalal, S., and Jansen, C. (2014). "MissMech: An R Package for Testing Homoscedasticity, Multivariate Normality, and Missing Completely at Random (MCAR)," Journal of Statistical Software, 56(6), 1-31. URL http://www.jstatsoft.org/v56/i06/.